---
title: "Mini Data-Analysis Deliverable 1"
output:github_doucment
    df_print: paged
---

# Welcome to your (maybe) first-ever data analysis project!

And hopefully the first of many. Let's get started:

1.  Install the [`datateachr`](https://github.com/UBC-MDS/datateachr) package by typing the following into your **R terminal**:

<!-- -->

    install.packages("devtools")
    devtools::install_github("UBC-MDS/datateachr")

2.  Load the packages below.

```{r}
library(datateachr)
library(tidyverse)
```

3.  Make a repository in the <https://github.com/stat545ubc-2022> Organization. You will be working with this repository for the entire data analysis project. You can either make it public, or make it private and add the TA's and Lucy as collaborators.

# Instructions

## For Both Milestones

-   Each milestone is worth 45 points. The number of points allocated to each task will be annotated within each deliverable. Tasks that are more challenging will often be allocated more points.

-   10 points will be allocated to the reproducibility, cleanliness, and coherence of the overall analysis. While the two milestones will be submitted as independent deliverables, the analysis itself is a continuum - think of it as two chapters to a story. Each chapter, or in this case, portion of your analysis, should be easily followed through by someone unfamiliar with the content. [Here](https://swcarpentry.github.io/r-novice-inflammation/06-best-practices-R/) is a good resource for what constitutes "good code". Learning good coding practices early in your career will save you hassle later on!

## For Milestone 1

**To complete this milestone**, edit [this very `.Rmd` file](https://raw.githubusercontent.com/UBC-STAT/stat545.stat.ubc.ca/master/content/mini-project/mini-project-1.Rmd) directly. Fill in the sections that are tagged with `<!--- start your work below --->`.

**To submit this milestone**, make sure to knit this `.Rmd` file to an `.md` file by changing the YAML output settings from `output: html_document` to `output: github_document`. Commit and push all of your work to the mini-analysis GitHub repository you made earlier, and tag a release on GitHub. Then, submit a link to your tagged release on canvas.

**Points**: This milestone is worth 45 points: 43 for your analysis, 1 point for having your Milestone 1 document knit error-free, and 1 point for tagging your release on Github.

# Learning Objectives

By the end of this milestone, you should:

-   Become familiar with your dataset of choosing
-   Select 4 questions that you would like to answer with your data
-   Generate a reproducible and clear report using R Markdown
-   Become familiar with manipulating and summarizing your data in tibbles using `dplyr`, with a research question in mind.

# Task 1: Choose your favorite dataset (10 points)

The `datateachr` package by Hayley Boyce and Jordan Bourak currently composed of 7 semi-tidy datasets for educational purposes. Here is a brief description of each dataset:

-   *apt_buildings*: Acquired courtesy of The City of Toronto's Open Data Portal. It currently has 3455 rows and 37 columns.

-   *building_permits*: Acquired courtesy of The City of Vancouver's Open Data Portal. It currently has 20680 rows and 14 columns.

-   *cancer_sample*: Acquired courtesy of UCI Machine Learning Repository. It currently has 569 rows and 32 columns.

-   *flow_sample*: Acquired courtesy of The Government of Canada's Historical Hydrometric Database. It currently has 218 rows and 7 columns.

-   *parking_meters*: Acquired courtesy of The City of Vancouver's Open Data Portal. It currently has 10032 rows and 22 columns.

-   *steam_games*: Acquired courtesy of Kaggle. It currently has 40833 rows and 21 columns.

-   *vancouver_trees*: Acquired courtesy of The City of Vancouver's Open Data Portal. It currently has 146611 rows and 20 columns.

**Things to keep in mind**

-   We hope that this project will serve as practice for carrying our your own *independent* data analysis. Remember to comment your code, be explicit about what you are doing, and write notes in this markdown document when you feel that context is required. As you advance in the project, prompts and hints to do this will be diminished - it'll be up to you!

-   Before choosing a dataset, you should always keep in mind **your goal**, or in other ways, *what you wish to achieve with this data*. This mini data-analysis project focuses on *data wrangling*, *tidying*, and *visualization*. In short, it's a way for you to get your feet wet with exploring data on your own.

And that is exactly the first thing that you will do!

1.1 Out of the 7 datasets available in the `datateachr` package, choose **4** that appeal to you based on their description. Write your choices below:

**Note**: We encourage you to use the ones in the `datateachr` package, but if you have a dataset that you'd really like to use, you can include it here. But, please check with a member of the teaching team to see whether the dataset is of appropriate complexity. Also, include a **brief** description of the dataset here to help the teaching team understand your data.

<!-------------------------- Start your work below ---------------------------->

1: CHOICE 1: cancer_sample 
2: CHOICE 2: steam_games
3: CHOICE 3: flow_sample
4: CHOICE_4: Vancouver_trees

<!----------------------------------------------------------------------------->

1.2 One way to narrowing down your selection is to *explore* the datasets. Use your knowledge of dplyr to find out at least *3* attributes about each of these datasets (an attribute is something such as number of rows, variables, class type...). The goal here is to have an idea of *what the data looks like*.

*Hint:* This is one of those times when you should think about the cleanliness of your analysis. I added a single code chunk for you below, but do you want to use more than one? Would you like to write more comments outside of the code chunk?

<!-------------------------- Start your work below 
For this project, I am first deciding the first four datasets to explore my interest as:
1.cancer_sample
2.vancouver_trees
3.flow_sample
4.buildings_permits
---------------------------->

```{r}
### EXPLORE HERE ###
datateachr::cancer_sample
head(cancer_sample)
class(cancer_sample)
summary(cancer_sample)

datateachr::flow_sample
head(flow_sample)
class(flow_sample)
summary(flow_sample)

datateachr::vancouver_trees
head(vancouver_trees)
class(vancouver_trees)
summary(vancouver_trees)


datateachr::building_permits
head(building_permits)
class(building_permits)
summary(building_permits)
```


`<!----------------------------------------------------------------------------->

1.3 Now that you've explored the 4 data sets that you were initially most interested in, let's narrow it down to 2. What lead you to choose these 2? Briefly explain your choices below, and feel free to include any code in your explanation.

<!-------------------------- Start your work below 
My choice is vancouver_trees in which I am interested as I love trees and the topic itself intrudes me, there are also variables of curb which has binary response and will be useful to run logistic regression.
I also want to work with the  cancer_sample data sets, I chose this  because I want to work with occupancy survey in future for my PhD thesis, whose response is 0 or 1 (binary categorical). 
.---------------------------->
 

<!----------------------------------------------------------------------------->

1.4 Time for the final decision! Going back to the beginning, it's important to have an *end goal* in mind. For example, if I had chosen the `titanic` dataset for my project, I might've wanted to explore the relationship between survival and other variables. Try to think of 1 research question that you would want to answer with each dataset. Note them down below, and make your final choice based on what seems more interesting to you!

<!-------------------------- Start your work below

For the vancouver_trees dataset, I want to explore if the height of the trees have any relationship with curb used.
For next dataset, I want to explore the relationship between diagnosis in which I want to explore if the diagnosis of cancer with being B or M is determined by area where, diagnosis as dependent (categorical) variable and area as independent variable.---------------------------->
 
<!---------------------------------------------------------------------------->

# Important note

Read Tasks 2 and 3 *fully* before starting to complete either of them. Probably also a good point to grab a coffee to get ready for the fun part!

This project is semi-guided, but meant to be *independent*. For this reason, you will complete tasks 2 and 3 below (under the **START HERE** mark) as if you were writing your own exploratory data analysis report, and this guidance never existed! Feel free to add a brief introduction section to your project, format the document with markdown syntax as you deem appropriate, and structure the analysis as you deem appropriate. Remember, marks will be awarded for completion of the 4 tasks, but 10 points of the whole project are allocated to a reproducible and clean analysis. If you feel lost, you can find a sample data analysis [here](https://www.kaggle.com/headsortails/tidy-titarnic) to have a better idea. However, bear in mind that it is **just an example** and you will not be required to have that level of complexity in your project.

# Task 2: Exploring your dataset (15 points)

If we rewind and go back to the learning objectives, you'll see that by the end of this deliverable, you should have formulated *4* research questions about your data that you may want to answer during your project. However, it may be handy to do some more exploration on your dataset of choice before creating these questions - by looking at the data, you may get more ideas. **Before you start this task, read all instructions carefully until you reach START HERE under Task 3**.

2.1 Complete *4 out of the following 8 exercises* to dive deeper into your data. All datasets are different and therefore, not all of these tasks may make sense for your data - which is why you should only answer *4*. Use *dplyr* and *ggplot*.

1.  Plot the distribution of a numeric variable.
2.  Create a new variable based on other variables in your data (only if it makes sense)
3.  Investigate how many missing values there are per variable. Can you find a way to plot this?
4.  Explore the relationship between 2 variables in a plot.
5.  Filter observations in your data according to your own criteria. Think of what you'd like to explore - again, if this was the `titanic` dataset, I may want to narrow my search down to passengers born in a particular year...
6.  Use a boxplot to look at the frequency of different observations within a single variable. You can do this for more than one variable if you wish!
7.  Make a new tibble with a subset of your data, with variables and observations that you are interested in exploring.
8.  Use a density plot to explore any of your variables (that are suitable for this type of plot).

2.2 For each of the 4 exercises that you complete, provide a *brief explanation* of why you chose that exercise in relation to your data (in other words, why does it make sense to do that?), and sufficient comments for a reader to understand your reasoning and code.

<!-------------------------- Start your work below 
1. I want to explore the variable height range and see if the data is highly skewed and see if there are any outliers in the data. 

```{r}
#First I want to understand the data and see the summary of the data set
vancouver_trees <- as.tibble(datateachr::vancouver_trees)
vancouver_trees$height_range_id <- as.factor(vancouver_trees$height_range_id)

#I want to print the distribution of the diameter in the data 
ggplot(vancouver_trees, aes(height_range_id)) +
  geom_bar(stata="identity", fill="dark green", color= "black")

```

2. Now, I want to explore how many missing data are there in the the curb variable and root barrier that I am interested in. This is also important as I may have to decide to deal with the missing values in required way. If you are replicating this project, don't miss doing it!

```{r}
#Check if there are many missing values 
any(is.na(vancouver_trees$curb))
any(is.na(vancouver_trees$root_barrier))

```
There are no missing data in the two predictor variables that I am interested in. Hence I am not worried now.

3. I want to conduct a regression and see if height of the trees is affected by the curb and the root barrier. But I am not sure if curb and root barrier are correlated to each other. If so, I may have to reconsider thinking of choosing only one of them to avoid collinearity. Since both variables are categorical that too nominal, I am conducting a chi square test.
```{r}
test <-chisq.test(vancouver_trees$curb, vancouver_trees$root_barrier)
test
ggplot(aes(curb), data = vancouver_trees) + 
  geom_bar()
  ggplot(aes(root_barrier), data = vancouver_trees)+ 
  geom_bar()
  
```
Since, both the value of p is 0.00408 which is less than 0.05, thus it is seen that the two explanatory variables are related to each other and I may have to omit one of them. I decide to omit curb and keep root_barriers. I have plotted the number of trees with root barrier and curb in 2 different plots.
4. I want to create subset of my the genus dataset to see if there is any relation of height and root barrier of Acer genus and Quercus species as the roots of Acer trees can spread out 12 feet wide and upto 3 feet deep and the root of Quercus goes 3-7 times deeper tha  their part above soil.
```{r}
datateachr::vancouver_trees
#Creating subset of Acer Genus
Acer_subset <- filter(vancouver_trees, genus_name == "ACER")
#Creating subset of Quercus genus
Quercus_subset <- filter(vancouver_trees, genus_name == "QUERCUS")
```

---------------------------------------------------------->
# Task 3: Write your research questions (5 points)

So far, you have chosen a dataset and gotten familiar with it through exploring the data. Now it's time to figure out 4 research questions that you would like to answer with your data! Write the 4 questions and any additional comments at the end of this deliverable. These questions are not necessarily set in stone - TAs will review them and give you feedback; therefore, you may choose to pursue them as they are for the rest of the project, or make modifications!

<!--- 
Question 1: How does diameter differ in two genus: Acer and Quercus?
Question 2: What are the diameters of the tree that are out of the range of most of the data in quercus genus?
Question 3: Is the diameter of Quercus species significantly affected by curb?
Question 4: Is the diameter of one height class significantly different than other height classes? --->


# Task 4: Process and summarize your data (13 points)

From Task 2, you should have an idea of the basic structure of your dataset (e.g. number of rows and columns, class types, etc.). Here, we will start investigating your data more in-depth using various data manipulation functions.

### 1.1 (10 points)

Now, for each of your four research questions, choose one task from options 1-4 (summarizing), and one other task from 4-8 (graphing). You should have 2 tasks done for each research question (8 total). Make sure it makes sense to do them! (e.g. don't use a numerical variables for a task that needs a categorical variable.). Comment on why each task helps (or doesn't!) answer the corresponding research question.

Ensure that the output of each operation is printed!

**Summarizing:**

1.  Compute the *range*, *mean*, and *two other summary statistics* of **one numerical variable** across the groups of **one categorical variable** from your data.
2.  Compute the number of observations for at least one of your categorical variables. Do not use the function `table()`!
3.  Create a categorical variable with 3 or more groups from an existing numerical variable. You can use this new variable in the other tasks! *An example: age in years into "child, teen, adult, senior".*
4.  Based on two categorical variables, calculate two summary statistics of your choosing.

**Graphing:**

5.  Create a graph out of summarized variables that has at least two geom layers.
6.  Create a graph of your choosing, make one of the axes logarithmic, and format the axes labels so that they are "pretty" or easier to read.
7.  Make a graph where it makes sense to customize the alpha transparency.
8.  Create 3 histograms out of summarized variables, with each histogram having different sized bins. Pick the "best" one and explain why it is the best.


Make sure it's clear what research question you are doing each operation for!

<!------------------------- Now I want to study the diameter differences between two genus. With the help of summary data, I will try to visualize. 

```{r}
#Question1: 
#Task 1: summary
summary(Acer_subset$diameter)
#300 diameter looks like outliers so to remove the outlier, lets create new #subset whose diameter is less than 100
Acer_subsetnew <- filter(Acer_subset, diameter < 100)
summary(Acer_subsetnew$diameter)
summary(Quercus_subset$diameter)
```
Now to graph and make histogram of diameter of Acer genus, we need to decide number of bins, we already know that the minimum diameter is 0 and maximum is 305. Now, since the gap is high, we need more bins. But let's try with small number of bins (just to see what happens!) 
```{r}
#Task 2: graphing with less number of bins
ggplot(Acer_subsetnew, aes(x=diameter))+
  geom_histogram(bins = 6)



```
We see that the information is not clear so lets use the formula we know <number of bins=(max-min)/width> 
```{r}
ggplot(Acer_subsetnew, aes(x=diameter))+
  geom_histogram(bins = 10)
```
That didn't work either, lets try with Struge's rule: log2(3607) + 1 = 13
```{r}

ggplot(Acer_subsetnew, aes(x=diameter))+
  geom_histogram(bins= 13)
```
Let's do next now!
```{r}
#QUESTION NUMBER 2
#Task 1: summarize
summary(Quercus_subset$diameter)
```
When we summarized the diameter of quercus, we could see diamter of 300 which is impossible! So certainly there is some outlier! Let's explore them!

```{r}
#Question number 2
#Task 2: visualizing
ggplot(data=Quercus_subset, aes(x=curb, y=diameter)) + geom_boxplot() + geom_count()
```
Can you see that one point which is completely out of range? That is outlier and needs to be removed.
Now let's do next question: Is the diameter of Acer significantly affected by curb? Let us set a null hypothesis that the mean diameter of tress  of having curb and not having curb is same.  
```{r}
#Question 3
#Task 1: summarize
t.test(Quercus_subset$diameter~Quercus_subset$curb, var.equal = FALSE)




```

Since the p-value of our test (2.2e-16) is less than alpha=0.05, we reject the null hypothesis of the test. This means we have sufficient evidence to say that the mean diameter of Quercus trees between having curb and not having curb is different.
```{r}
#Task 2: graph

#ggplot(Quercus_subset, aes(Quercus_subset$curb, Quercus_subset$diameter) + #geom_boxplot(outlier.color = "red", outlier.shape = 8, outlier.size = 4))
#Alternatively, you can also use the following commands 
boxplot(Quercus_subset$diameter~Quercus_subset$curb)

```
We can see that the average diameter of the Quercus not having curbs is slightly more than the ones which have curb.
Now lets move to the 4th question: Since the height is given as categorical variable and y variable is diameter which is continuous variable, linear regression model is to be done:
```{r}
#Question 4
#Task1: summary
?lm()
model = lm(vancouver_trees$diameter~vancouver_trees$height_range_id)
summary(model)
```
From this we can see that the relationship of diameters is significant to each height range class. 

### 1.2 (3 points)
```{r}
#question4: task 2: graphing
# I want to group the diameter according to height class 
#And plot graph which shows height class in x axis and average diameter of the class for each class in y axis
groupeddiameter <- vancouver_trees %>%
group_by(height_range_id) %>%
summarize(a = mean(diameter))
ggplot(data= groupeddiameter, aes(x= height_range_id, y= a)) +
geom_bar(stat= "identity")

  


```
We can see that the diameter is increasing with increase in height classes in general except for the last height class. 

---------------------------------------------->
Based on the operations that you've completed, how much closer are you to answering your research questions? Think about what aspects of your research questions remain unclear. Can your research questions be refined, now that you've investigated your data a bit more? Which research questions are yielding interesting results?

<!-------------------------- Start your work below
I am very close to my research questions. I tried to find summary and visualize in order to deal with the research questions. 

The interesting results are that the average diameter of the Quercus not having curbs is slightly more than the ones which have curb. Hence curbs are not good for plant growth for Quercus trees. 
---------------------------->
<!----------------------------------------------------------------------------->

### Attribution

Thanks to Icíar Fernández Boyano for mostly putting this together, and Vincenzo Coia for launching.
